---
title: "ANOVA_activity"
author: "Mallikarjuna Mannem"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install packages

```{r}
# Load necessary packages
library(tidyverse)
library(ggthemes)
library(flextable)
library(broom)
library(ggfortify)

```

```{r}
# Set ggplot theme for visualizations
theme_set(ggthemes::theme_few())

# Set options for flextables
set_flextable_defaults(na_str = "NA")

# Load function for printing tables nicely
source("https://raw.githubusercontent.com/dilernia/STA323/main/Functions/make_flex.R")
```

## GPA example

```{r}
# Importing fake GPA data
simGPA <- readr::read_csv("https://raw.githubusercontent.com/dilernia/STA323/main/Data/simGPA.csv")

# Creating vector of colors
schoolColors <- setNames(c("#4E2583", "#5470c0", "#aa152b"),
                         c("AU", "GVSU", "SVSU"))
set.seed(1994)

# Printing random rows of data set
simGPA %>% 
  dplyr::filter(University %in% c("GVSU", "SVSU")) %>% 
  slice_sample(n = 7) %>% 
  make_flex()
```

We can create side-by-side box plots to compare the GPAs of two the
universities

```{r}
# Creating side-by-side box plots
twoSampGG <- simGPA %>% 
  dplyr::filter(University %in% c("GVSU", "SVSU")) %>% 
  ggplot(aes(x = University, y = GPA, fill = University)) + 
  geom_boxplot() + 
scale_fill_manual(values = schoolColors) + 
    theme(legend.position = "none")
  
twoSampGG
```

we can also calculate summary statistics for each university

```{r}
# Calculating descriptive statistics
quantStats <- simGPA %>% 
  dplyr::filter(University %in% c("GVSU", "SVSU")) %>% 
  group_by(University) %>% 
  summarize(
  Minimum = min(GPA, na.rm = TRUE),
  Q1 = quantile(GPA, na.rm = TRUE, probs = 0.25),
  M = median(GPA, na.rm = TRUE),
  Q3 = quantile(GPA, na.rm = TRUE, probs = 0.75),
  Maximum = max(GPA, na.rm = TRUE),
  Mean = mean(GPA, na.rm = TRUE),
  R = Maximum - Minimum,
  s = sd(GPA, na.rm = TRUE),
  n = n()
)

# Printing table of statistics
quantStats %>% 
  make_flex(caption = "Summary statistics for student GPAs by university.")
```

we consider the case in which we have a third university we want to
compare.

```{r}
# Creating side-by-side box plots
threeSampGG <- simGPA %>% 
  ggplot(aes(x = University, y = GPA, fill = University)) + 
  geom_boxplot() + 
scale_fill_manual(values = schoolColors) + 
    theme(legend.position = "none")
  
threeSampGG
```

we can calculate summary statistics

```{r}
# Calculating descriptive statistics
quant3Stats <- simGPA %>% 
  group_by(University) %>% 
  summarize(
  Minimum = min(GPA, na.rm = TRUE),
  Q1 = quantile(GPA, na.rm = TRUE, probs = 0.25),
  M = median(GPA, na.rm = TRUE),
  Q3 = quantile(GPA, na.rm = TRUE, probs = 0.75),
  Maximum = max(GPA, na.rm = TRUE),
  Mean = mean(GPA, na.rm = TRUE),
  R = Maximum - Minimum,
  s = sd(GPA, na.rm = TRUE),
  n = n()
)

# Printing table of statistics
quant3Stats %>% 
  make_flex(caption = "Summary statistics for student GPAs by university.")
```

## survival example

let's first import the data into R

```{r}
# Loading survival data
cancerSurvival <- readr::read_csv("https://raw.githubusercontent.com/dilernia/STA323/main/Data/CancerSurvival.csv")
```

printing 8 randomly selected rows from the data set

```{r}
set.seed(1994)

# Printing random rows of data set
cancerSurvival %>%
  slice_sample(n=8) %>%
  make_flex()
```

```{r}
# Creating side-by-side box plots
cancerSurvival %>% 
  ggplot(aes(x = Organ, y = Survival, fill = Organ)) + 
  geom_boxplot() + 
  labs(y = "Survival time (days)",
       title = "Duration of survival from time of diagnosis") +
  scale_fill_viridis_d() +
    theme(legend.position = "none")
```

Next, we calculate summary statistics for the survival of each cancer
type

```{r}
# Calculating descriptive statistics
quant3Stats <- cancerSurvival %>% 
  group_by(Organ) %>% 
  summarize(
  Minimum = min(Survival, na.rm = TRUE),
  Q1 = quantile(Survival, na.rm = TRUE, probs = 0.25),
  M = median(Survival, na.rm = TRUE),
  Q3 = quantile(Survival, na.rm = TRUE, probs = 0.75),
  Maximum = max(Survival, na.rm = TRUE),
  Mean = mean(Survival, na.rm = TRUE),
  R = Maximum - Minimum,
  s = sd(Survival, na.rm = TRUE),
  n = n()
)

# Printing table of statistics
quant3Stats %>% 
  make_flex(caption = "Summary statistics for survival times in days by cancer type.")
```

**Based on the output obtained, which cancer type has the longest
survival on average and what is the corresponding average survival
time?**

Breast cancer,1395.91

**Which cancer type has the most variability in its survival times?
Why?**

breast cancer since R-value and s-value are more for breast cancer.

**What was the shortest duration of survival for any individual
person,and what was the longest? Specify which type of cancer each
person had.**

Shortest duration of survival is for the individuals with 20 days with
Bronchus and Colon cancer. Longest is for the individual with breast
cancer and survival duration of 3808 days.

**Provide a statement of the group means formulation of the One-Way
ANOVA model.**

The group effects formulation of ANOVA model in general is:
$$Y_{ij}=\mu_i+\varepsilon_{ij}$$ Response=Grand Mean +Group Effect
+Error Term $Y_{ij}$ : is the survival time in days for the jth
observation of the ith group \$\mu\_i : mean survival time of ith cancer
type $\varepsilon_{ij}$ : independent and normally distributed error
terms

###Hypotheses

Our hypotheses are given by

Ho: μ1 = μ2 = μ3 = μ4 = μ5 vs. Ha: μi ≠ μj for some i,j

where μ1 is the average duration of survival in days for breast cancer
patients, μ2 for bronchus cancer, μ3 for colon cancer, μ4 for ovarian
cancer, and μ5 for stomach cancer.

Alternative statement of hypotheses:

Our hypotheses are given by

Ho: μ1 = μ2 = μ3 = μ4 = μ5 vs. Ha: μi ≠ μj, for some i,j

where μi is the average duration of survival in days for the ith cancer
type for i=1,2,...,5.

**What are the response and explanatory variables for this One-Way
ANOVA?**

r: survival time in days e: cancer type

```{r}
# Fitting One-Way ANOVA model
survivalModel <- aov(Survival ~ Organ, data = cancerSurvival)

# Printing ANOVA table
survivalModel %>% 
  broom::tidy() %>% 
  make_flex(caption = "ANOVA table for survival duration in days by cancer type")
```

-   SSGroups = 11,535,760.52, this is the total variability in the
    survival times of the patients explained by their cancer type

-   SSE = 26,448,144.48, this is the total variability in the survival
    times of the patients not explained by their cancer type

### Assumptions

1.  **Have the same standard deviation within each group(implies equal
    variances, also known as homoskedasticity)**

```{r}
autoplot(survivalModel, which = 1, label.repel = TRUE)@plots[[1]]
```

There is an outward fanning pattern in the residual by fitted value
plot, indicating that the homoskedasticity assumption is violated.

**Ratio of standard deviations**

Smax = 1238.97

Smin = 209.86

1238.97 / 209.86 = 5.90 which is greater than 2, so this indicates that
the homoskedasticity assumption is violated.

2.  **Be normally distributed**

**QQ plot**

```{r}
# Quantile-quantile (QQ) plot
autoplot(survivalModel, which = 2, label.repel = TRUE)@plots[[1]]

```

There are many points on the QQ plot that stray far from the diagonal
line through the origin, indicating that the normality assumption is
violated.

**Histogram of the residuals**

```{r}
# Histogram of the residuals
survivalModel %>% 
  broom::augment() %>% 
  ggplot(aes(x = .resid)) + 
  geom_histogram(color = "white") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.10))) +
  labs(title = "Distribution of residuals",
       x = "Residual",
       y = "Frequency")
```

The distribution is right-skewed, indicating that the normality
assumption is violated.

3.  **Be independent**

since these are measurements for seperate individuals, we will consider
this assumption to be sufficiently met.

The assumptions were violated, so truly the results of the ANOVA model
are unavailable.

However, for demonstration purpose we will finish conducting the overall
F-test.

**Provide the value of the test statistic, p-value, and decision for the overall F-test, citing specific evidence from the obtained output.**

**Test statistic**: F = 6.43

**p-value**: 0.00023

**Decision**: Reject Ho since the p-value is less than 0.05

**Provide our interpretation based on our stated decision in the context of the problem.**

Since the p-value is less than 0.05, we have sufficient evidence that at
least two cancer types differ in their average survival time in days at
the 5% significance level.

```{r}
# Calculating log-transformation of the survival times
cancerSurvival <- cancerSurvival %>% 
  dplyr::mutate(logSurvival = log(Survival))
# Creating side-by-side box plots
cancerSurvival %>% 
    ggplot(aes(x = Organ, y = logSurvival, fill = Organ)) + 
    geom_boxplot() + 
    labs(y = "Log survival time (days)",
         title = "Duration of survival from time of diagnosis") +
    scale_fill_viridis_d() +
    theme(legend.position = "none")
```

*Fit a One-Way ANOVA model using a log transformation of the survival times*

```{r}
# Fitting One-Way ANOVA model
logSurvivalModel <- aov(logSurvival ~ Organ, data = cancerSurvival)

# Printing ANOVA table
logSurvivalModel %>% 
  broom::tidy() %>% 
  make_flex(caption = "ANOVA table for log survival duration in days by cancer type")
```

**Checking assumptions for the log transformation model**

```{r}
# Calculating descriptive statistics
quantStats <- cancerSurvival %>% 
  group_by(Organ) %>% 
  summarize(
  Minimum = min(logSurvival, na.rm = TRUE),
  Q1 = quantile(logSurvival, na.rm = TRUE, probs = 0.25),
  M = median(logSurvival, na.rm = TRUE),
  Q3 = quantile(logSurvival, na.rm = TRUE, probs = 0.75),
  Maximum = max(logSurvival, na.rm = TRUE),
  Mean = mean(logSurvival, na.rm = TRUE),
  R = Maximum - Minimum,
  s = sd(logSurvival, na.rm = TRUE),
  n = n()
)

# Printing table of statistics
quantStats %>% 
  make_flex(caption = "Summary statistics for log duration of survival from time of diagnosis by cancer type.")
```

## Checking assumptions

**Homoskedasticity**

since 1.65/0.95 = 1.736842 is less than 2, and since the vertical spread
is fairly constant in the residual by fitted value plot, the
homoskedasticity assumption is met for the log-transformed model.

**normality**

since the points in the QQ plot do not stray far from the diagonal line,
this indicates that the normality assumuption is met for the
log-transformed model.

Moreover, since the histogram unimodal and only very slightly
left-skewed, this together with the QQ plot indicates that the normality
assumption is met for the log-transformed model.

Overall, the assumptions are now met, so the results of the ANOVA model
for the log-transformed values are reliable.

**Test statistic**: F = 4.29

**p-value**: 0.0041

**Decision**: Reject Ho since the p-value is less than 0.05

##Provide our interpretation based on our stated decision in the context
of the problem.

Since the p-value is less than 0.05, we have sufficient evidence that at
least two cancer types differ in their average survival time in days at
the 5% significance level.

## Post-Hoc Analysis

**Provide and interpret the confidence interval for the average log
survival of patients with breast cancer based on the log transformed
output**

confidence limits: (5.84,7.28)

We are 95% confident that the mean log survival time in days for breast
cancer patients is between 5.84 and 7.28

**A better interpretation after having used a log-transformation**

We are 95% confident that the mean survival time in days for breast
cancer patients is between 344 and 1451 days.

Next we obtain post-havoc output to compare pairs of groups

```{r}
# Conducting post-hoc analysis using Tukey's HSD for pairwise differences
pairwisePostHoc <- TukeyHSD(logSurvivalModel)

# Printing post-hoc output
pairwisePostHoc %>% 
  broom::tidy() %>% 
  make_flex(caption = "Pairwise 95% confidence intervals for differences between cancer types in terms of average log survival in days.")
```

*From row 1):*

we have sufficient evidence that breast cancer patients have a longer
log survival time in days on average, than bronchus cancer patients.

c limits: (-2.91, -0.30)

we are 95% confident that the mean log duration of survival in days for
bronchus cancer patients is between 0.30 and 2.91 less than that of
breast cancer patients

*From row 4):*

we have sufficient evidence that breast cancer patients have a longer
log survival time in days on average, than stomach cancer patients.

C limits: (-2.97, -0.21)

we are 95% confident that the mean log duration of survival in days for
stomach cancer patients is between 0.21 and 2.97 less than that of
breast cancer patients.

**we will not have to do this back-transformation on exam**

**From row 1):**

we have sufficient evidence that breast cancer patients have a longer
survival time in days on average, than bronchus cancer patients.

ca limits: log-scale (2.91, 0.30) ---\> ---\> (18.356799, 1.349859)

we are 95% confident that the mean duration of survival in days for
breast cancer patients is between 1.3 and 18.3 times/ that of bronchus
cancer patients.
