---
title: "Categorical Predictors & Dummy Variables"
author: "Arjun Mannem"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

```{r}
# Load necessary packages
library(tidyverse)
library(ggthemes)
library(flextable)
library(corrr)
library(ggfortify)
library(broom)
library(car)
```

```{r}
# Set ggplot theme for visualizations
theme_set(ggthemes::theme_few())

# Set default ggplot2 colors
my_palette <- ggthemes::colorblind_pal()(8)[c(4, 6:8, 2:3, 5, 1)]

# Set options for flextables
set_flextable_defaults(na_str = "NA")

# Load function for printing tables nicely
source("https://raw.githubusercontent.com/dilernia/STA323/main/Functions/make_flex.R")

# Load function for visualizing VIF & GVIF values
source("https://raw.githubusercontent.com/dilernia/STA323/main/Functions/vif_plot.R")
```



```{r}
# Loading penguins data
penguins <- readr::read_csv("https://raw.githubusercontent.com/dilernia/STA323/main/Data/penguins.csv")

# Dropping missing values
penguinsComplete <- penguins %>% 
  dplyr::filter(!is.na(sex)) %>% 
    dplyr::mutate(sex = fct_relevel(sex, "male", "female"))

# Printing sample of rows from data set
penguinsComplete %>% 
  dplyr::select(species, sex, body_mass_g, flipper_length_mm) %>% 
  dplyr::sample_n(size = 5) %>% 
  make_flex(caption = "Five randomly selected rows and four variables from the Palmer Penguins data.")
```



```{r}
# Creating a scatter plot of body mass by flipper length
penguinsComplete %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(pch = 21, fill = "black",
             color = "white") + 
  geom_smooth(aes(x = flipper_length_mm, y = body_mass_g), method = "lm", se = FALSE) +
  labs(y = "Body mass (g)", 
       x = "Flipper length (mm)", 
       title = "Palmer penguins data: body mass by flipper length")
```


```{r}
# Fitting SLR model
slrModel <- lm(body_mass_g ~ flipper_length_mm, data = penguinsComplete)

# Printing SLR model output
slrModel %>% 
  broom::tidy() %>% 
  make_flex(ndigits = 1,
            caption = "SLR model output.")
```


**How do we interpret the estimated slope?**

For every 1mm increase in a penguin's flipper length, we expect its body mass to increase by 50.2 grams.


**How do we interpret the estimated intercept?**

Since 0 is outside the scope of the data set


#### What if we wanted to incorporate categorical variables in our model?

```{r}
# Creating scatter plot of body mass by flipper length coloring points by sex
penguinsComplete %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(aes(x = flipper_length_mm, y = body_mass_g, fill = sex), pch = 21, color = "white") + 
  geom_smooth(aes(x = flipper_length_mm, y = body_mass_g), 
              method = "lm", se = FALSE, color = "black") +
  labs(y = "Body mass (g)", x = "Flipper length (mm)", title = "Palmer penguins data: body mass by flipper length") +
    scale_fill_manual(values = my_palette) +
  theme(legend.position = "bottom")
```


```{r}
# Creating side-by-side boxplots for body mass by sex
penguinsComplete %>% 
  ggplot(aes(x = sex, y = body_mass_g, fill = sex)) + 
  geom_boxplot() + 
  labs(y = "Body mass (g)", x = "sex", title = "Palmer penguins data: body mass by sex") +
  scale_fill_manual(values = my_palette) + 
  theme(legend.position = "none")
```



**Does a penguin’s body mass (g) appear to vary based on its sex? Do male or female penguins have larger average body masses?**




### Categorical Predictors

Example 1: Predicting how many points a basketball team will score using playing location (home, away, or a neutral court)

Example 2: Estimating the effect of different treatments on blood pressure (Drug A, Drug B, and a placebo)


**How many dummy variables would be needed to incorporate the categorical predictors in Examples 1 and 2?**

Example 1: 2 dummy variables

Example 2: 2 dummy variables

### Palmer Penguins Data

```{r}
# Set seed value for reproducible random number generation
set.seed(1994)

# Creating explicit dummy variable for sex of the penguin
penguinsDummy <- penguinsComplete %>% 
  dplyr::mutate(sexfemale = ifelse(sex == "female", 1, ifelse(sex == "male", 0, NA)))

# Displaying data with dummy variable
penguinsDummy %>% 
  dplyr::select(species, sex, sexfemale, body_mass_g, flipper_length_mm) %>% 
  sample_n(size = 5) %>%
  make_flex(caption = "Five randomly selected rows from the Palmer Penguins data displaying a dummy variable.")
```


#### Modeling Penguin Body Mass: Theoretical Model

```{r}
# Group means for plotting line of best fit
bodyMassMeans <- penguinsDummy %>% 
  dplyr::group_by(sex) %>% 
  dplyr::summarize(Mean = mean(body_mass_g), 
                   sexfemale = mean(sexfemale))

# Visualizing group means
penguinsDummy %>% 
  ggplot(aes(x = sexfemale, y = body_mass_g, 
             fill = sex)) + 
  geom_point(pch = 21, color = "white") +
  geom_hline(data = bodyMassMeans, 
              aes(yintercept = Mean, color = sex)) +
    labs(y = "Body mass (g)", x = "Female Indicator", title = "Palmer penguins data: body mass by sex") +
  scale_fill_manual(values = my_palette) + 
    scale_color_manual(values = my_palette) + 
  scale_x_continuous(breaks = c(0, 1)) +
  theme(legend.position = "bottom")
```

```{r}
# Fitting SLR model with sex as predictor
slrModel2 <- lm(body_mass_g ~ sex, 
                data = penguinsComplete)

# Printing SLR model output
slrModel2 %>% 
  broom::tidy() %>% 
  make_flex(ndigits = 1,
            caption = "SLR model output.")
```

**Which group is the baseline / reference category?**

The male category is the baseline / reference category since it does not have a dummy variable as a predictor in the model.

**How do we interpret the estimated slope?**

Since βHat1 = -683.4, we expect that female penguins have body masses that are 683.4 grams less than that of male penguins, on average.

**How do we interpret the estimated intercept?**

Since βHat0 = 4,545.7, this means that the expected body mass of the male penguins is 4,545.7 grams.

**What is the expected body mass (g) of a female penguin?**

yHat = 4545.7 − 683.4 * 1 = 3862.3 grams


#### Modeling Penguin Body Mass

```{r}
# Creating scatter plot of body mass by flipper length colored by sex
penguinsComplete %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g,
             fill = sex)) + 
  geom_point(pch = 21, color = "white") + 
  labs(y = "Body mass (g)", x = "Flipper length (mm)", title = "Palmer Penguins Data: Body Mass by Flipper Length") +
  scale_fill_manual(values = my_palette) + 
  theme(legend.position = "bottom")

```

```{r}
# Fitting MLR model
mlrModel <- lm(body_mass_g ~ sex + flipper_length_mm,
               data = penguinsComplete)

# Extracting MLR model coefficients
betaHats <- coefficients(mlrModel)

# Printing MLR model output
mlrModel %>% 
  broom::tidy() %>% 
  make_flex(ndigits = 1,
            caption = "MLR model output.")
```

**How do we interpret the estimated slope for the sex of the penguin?**

Since βHat1 = -347.9, we expect female penguins have body masses that are 347.9 grams less than that of male penguins with same flipper length.

**How do we interpret the estimated intercept?**

Since flipper length = 0 is outside the scope of the data set, interpreting the intercept would be extrapolation, and thus is inappropriate.

**How do we interpret the estimated slope for the flipper lengths of the penguins?**

Since βHat2 = 47.0, for every 1mm increase in a penguin's flipper length, we expect its body mass to increase by 47.0 grams.


#### Modeling Penguin Body Mass using Species

**How many dummy variables are used to represent the species variable?**

2 dummy variables since there are 3 species.

**Based on the model above, which species is the baseline / reference category? Why?**

Adelie species is the baseline / reference category,since it does not have a dummy variable.( X1=1 for an chinstrap penguin and 0 otherwise, X2=1 for a gentoo penguin and 0 otherwise.)


```{r}
# Set seed value for reproducible random number generation
set.seed(1994)

# Creating explicit dummy variables for species
penguinsDummy2 <- penguinsComplete %>% 
  dplyr::mutate(speciesChinstrap = ifelse(species == "Chinstrap", 1, 0),
         speciesGentoo = ifelse(species == "Gentoo", 1, 0))

# Displaying data with dummy variabels for species
penguinsDummy2 %>% 
  dplyr::select(species, speciesChinstrap, speciesGentoo, body_mass_g, flipper_length_mm) %>% 
  dplyr::sample_n(size = 5) %>%
  make_flex(caption = "Five randomly selected rows from the Palmer Penguins data displaying dummy variables.")

```


```{r}
# Fitting linear model with species as predictor
speciesModel <- lm(body_mass_g ~ species, 
                   data = penguinsComplete)

betaHats <- coefficients(speciesModel)
# Printing MLR model output
speciesModel %>% 
  broom::tidy() %>% 
  make_flex(ndigits = 1,
            caption = "MLR model output.")
```

yHat = 3706.2 + 26.9X1 + 1386.3X2

**Interpretation of estimated intercept**: Since βHat0 = 3706.2, the expected bodymass of an Adelie penguin is 3706.2 grams.

**Interpretation of estimated slope for Chinstrap penguins**: Since βHat1 = 26.9, the expected bodymass of an Chinstrap penguin is 
26.9 grams greater than that of Adelie species. Here, 3706.2 + 26.9 = 3733.1 grams.

**Interpretation of estimated slope for Chinstrap penguins**: Since βHat2 = 1386.3, the expected bodymass of an Chinstrap penguin is 
1386.3 grams greater than that of Adelie species. Here, 3706.2 + 1386.3 = 5092.5 grams.


```{r}
# Group means for plotting line of best fit
bodyMassMeans <- penguinsDummy %>% 
  dplyr::group_by(species) %>% 
  dplyr::summarize(Mean = mean(body_mass_g))

# Plotting group data
penguinsDummy %>% 
  ggplot(aes(x = as.integer(as.factor(species)), y = body_mass_g, 
             fill = species)) + 
  geom_point(pch = 21, color = "white") +
  labs(y = "Body mass (g)", x = "", title = "Palmer Penguins Data: Body Mass by species") +
   geom_hline(data = bodyMassMeans, 
              aes(yintercept = Mean, color = species)) +
  scale_fill_manual(values = my_palette[4:6]) +
  scale_color_manual(values = my_palette[4:6]) +
  scale_x_continuous(breaks = 0) +
  theme(legend.position = "bottom")
```

```{r}
# Fitting linear models
speciesFlipModel <- lm(body_mass_g ~ species + flipper_length_mm, data = penguinsComplete)

betaHats <- coefficients(speciesFlipModel)

# Printing MLR model output
speciesFlipModel %>% 
  broom::tidy() %>% 
  make_flex(ndigits = 1,
            caption = "MLR model output.")
```

**How do we interpret the estimated intercept?**

βHat0 = -4013.2

Since a flipper length of 0mm is outside the range of our observed data set, it would be **extrapolation** to interpret this value, and thus this is inappropriate.

**How do we interpret the estimated slopes?**

(1) Since βHat_Chinstrap = -205.4, the expected body mass in grams of Chinstrap penguins is 205.4 grams less than that of the Adelie penguins with the same flipper lengths.

(2) Since βHat_Gentoo = 284.5, the expected body mass in grams of Gentoo penguins is 284.5 grams more than that of the Adelie penguins with the same flipper lengths.

(3) Since βHat_Flipperlength = 40.6, for every 1mm increase in penguin's flipper length, we expect its body mass in grams to increase by 40.6 grams, holding its species constant.


**What is the expected body mass (g) of a Chinstrap penguin with a flipper length of 195mm? Is this extrapolation?**

yHat = -4013.2 - 205.4 * (x1) +284.5 * (x2) + 40.6 * (x3)

yHat = -4013.2 - 205.4 * (1) +284.5 * (0) + 40.6 * 195

yHat = 3698.4 grams

No, this is not extrapolation since 195mm is within the range of the observed flipper lengths for the Chinstrap penguins.

```{r}
# Obtaining predicted value
predict(speciesFlipModel, 
        newdata = tibble(species = "Chinstrap",
                         flipper_length_mm = 195))
##        1 
## 3699.648

```


**If a Chinstrap penguin with a flipper length of 195mm had a body mass of 3900 grams, what is its corresponding residual? What does this mean?**

residual = y - yHat

residual = 3900 - 3698.4 = 201.6 grams

This penguin weighed about slightly more(200 grams) than our model expected, given its species and flipper lengths.


**What is the expected body mass (g) of a Chinstrap penguin with a flipper length of 225mm? Is this extrapolation?**

Since 225mm is outside the range of the observed flipper lengths for the Chinstrap penguins, such a predicted value would be extrapolation, and thus is inappropriate to calculate.

**Does it seem reasonable to constrain the regression lines to be parallel for each group?**

Since parallel lines fit the data reasonably well, it does seem reasonable to constrain the regression lines to be parallel for each group.







