---
title: "Fundamentals of Functions in R"
author: "Arjun Mannem"
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
output:
  rmdformats::readthedown:
    css: gvsu_readthedown_theme.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE,error = TRUE)
```

**First, let’s load packages necessary for this activity using the code below. Note: you may need to install packages beforehand by using the install.packages() function and the package name in quotes.**

```{r}
library(tidyverse)
library(stringr)
library(lubridate)
```

## Example function

```{r}
# Creating a function called mySum with 2 required arguments
mySum <- function(number1, number2) {
  return(number1 + number2)
}

# Defaulting to order of arguments in function
mySum(5, 3)

# Manually specifying arguments
mySum(number1 = 5, number2 = 3)

mySum(TRUE,FALSE)
```

**Create the mySum() function, and use it to calculate the sum of 13 and 1989.**

```{r}
mySum(13,1989)
```

**What happens when we calculate the sum of NA and 1989?**

```{r}
mySum(NA,1989)
```
## You try

**Copy the code creating the myNewSum() function below and add comments to communicate the purpose of different parts of the code.**

```{r}
# Function to calcuulate sum of two numbers and possibly remove the missing values.
## number1: a single numeric value
## number2: a single numeric value
## remove.na: whether to remove missing values (TRUE) or not (FALSE)

myNewSum <- function(number1, number2, remove.na = TRUE) {
  if(remove.na) {
    calcSum <- ifelse(is.na(number1), 0, number1) + 
         ifelse(is.na(number2), 0, number2)
  } else {
    calcSum <- number1 + number2
  }
  return(calcSum)
}
```

**Read the source code for each of the following three functions, determine the purpose of each function, and propose better function names.**

```{r}
# Function to check if the prefix matches
check_prefix <- function(string, prefix) {
  str_sub(string, 1, nchar(prefix)) == prefix
}

check_prefix("anti-establishment","anti")


# Function to drop last value in a vector
drop_last <- function(my_vector) {
  if (length(my_vector) <= 1) return(NULL)
  else return(my_vector[-length(my_vector)])
}

drop_last(1:5)


# Harmonic Mean
harmonic_mean <- function(x) {
  return (1 / (mean(1 / x)))
}

harmonic_mean(1:5)
```

**Create a new function to calculate the geometric mean of a vector of numbers which is equal to the nth root of the product of n numbers.**

```{r}

geometric_mean <- function(values) {
  return(exp(mean(log(values))))
  #return ((prod(values)^(1/length(values)))) #gives Infinity for large number
}

geometric_mean(1:10)
```

## Conditional execution, warnings, and errors

```{r}
# Function to calculate square root
calcSqrt <- function(val) {
if (val >= 0) {
  return(val^0.5)
} else {
  warning("val is negative but must be >= 0!")  
  #stop("val is negative but must be >= 0!") 
  #message("If you need to drop a message")
  return(NaN)
}
}

calcSqrt(4)
## [1] 2
calcSqrt(-9)
## [1] NaN

```

```{r}

foodType <- function(food) {
if (food %in% c("apple", "orange", "banana")) {
  return("fruit")
} else if(food %in% c("broccoli", "asparagus")) {
  return("vegetable")
} else {
  return("other")
}}

foodType("Reese's Puffs")
foodType("orange")
```

**Create a function called is.whole.number(), which checks if a given value is a whole number (e.g., 0, 1, 2, …) or not.**

```{r}

is_whole_number <- function(x){
  if(x>=0 & (x %% 1)==0) return(TRUE)
  else return(FALSE)
  

  #efficient way
  ##if(x%%1 == 0) return(FALSE)
  ##if(x<0) return(FALSE)
  ##]else return(TRUE)
}

is_whole_number(-1)
is_whole_number(0)
is_whole_number(3)
is_whole_number(3.1)

```

**The ^ operator raises a number to a specified power in R. Create a function that raises a number to a specified power, and returns a warning and an NaN value if the number is negative and the power is not a whole number.**

```{r}
raise_power <- function(number, power){
  
}
```




## The ifelse() function

```{r}
# Creating a function called calculate_sum with 2 required arguments, 1 optional argument
calculate_sum <- function(number1, number2, remove_na = TRUE) {
  if(remove_na) {
    calcSum <- ifelse(is.na(number1), 0, number1) + 
         ifelse(is.na(number2), 0, number2)
  } else {
    calcSum <- number1 + number2
  }
  return(calcSum)
}

calculate_sum(13,1989)
```

What happens when we calculate the sum of NA and 1989 using calculate_sum()?

It replace NA by 0. result = 1989.

## Logical operators

```{r}
# The 'or' operator
(3 < 5) || (-1 > 0)

# The 'and' operator
(3 < 5) && (-1 > 0)
```

```{r}
# The 'equals' operator
3 == 5
(4/2) == 2

3 != 5
# Vectorized examples
1:4
1:4 == 3

(1:4 < 4) & (1:4 > 1)

(1:4 < 4) | (1:4 > 1)

```

Create a function that takes a numeric vector as input, and returns a single boolean value indicating whether or not the vector has any negative values.

```{r}
any_negatives1 <- function(vector) {
  any_negative <- sum(vector) != sum(abs(vector))
  return(any_negative)
}

any_negatives2 <- function(vector) {
  any_negative <- !near(sum(vector),sum(abs(vector)))
  return(any_negative)
}

any_negatives3 <- function(vector) {
  any_negative <- any(vector < 0)
  return(any_negative)
}

any_negatives1(c(1,0,pi,-1))
any_negatives2(c(1,0,pi,-1))
any_negatives3(c(1,0,pi,-1))
```

Does the code (1:4 < 4) & (1:4 > 1) produce the same result as (1:4 < 4) && (1:4 > 1)? Which code achieves what appears to be its intended purpose?

(1:4 < 4) & (1:4 > 1)

## You try

Create a function to check if a number is even or odd that returns a vector containing the values “even”, “odd”, or “not an integer” depending on its input.

```{r}
even_or_odd <- function(int_vector){
  
  return(ifelse(int_vector %% 1 != 0,"not an integer",
         ifelse(int_vector %% 2 == 0, "even","odd")))
  
}

even_or_odd(c(2,5,pi))
```

Create a function called print_greeting() that has a single argument, date, with a default value of today(), such that print_greeting() prints the following greeting when called:

```{r}
print_greeting <- function(user,date = today()){
  message <- paste0("Hello ",user,"! Today is ", 
                    wday(date, abbr = FALSE, label = TRUE),
                    ".")
  return(message)
}

print_greeting("Arjun")

```

Extend the print_greeting() function to include the date as well like in the message below. Hint: use scales::ordinal(lubridate::day(today())) to display the date with the ordinal (th or nd) included.

```{r}
library(purrr)

print_greeting <- function(user,date = today()){
  message <- paste0("Hello ",user,"! Today is ", 
                    wday(date, abbr = FALSE, label = TRUE),
                    ",", month(date, label = TRUE, abbr = FALSE),
                    scales::ordinal(lubridate::day(today())),
                    ".")
  return(message)
}

print_greeting("Arjun")

```


Further extend the print_greeting() function to have an optional argument, user, that customizes who the greeting is for and has a default value of Sys.info()[["user"]] as shown below.

```{r}
library(purrr)

print_greeting <- function(date = today(),user = Sys.info()[["user"]]){
  message <- paste0("Hello ",user,"! Today is ", 
                    wday(date, abbr = FALSE, label = TRUE),
                    ",", month(date, label = TRUE, abbr = FALSE),
                    scales::ordinal(lubridate::day(today())),
                    ".")
  return(message)
}

print_greeting(user = "Florence Nightingale")

```

Further extend the print_greeting() function to say “Good morning” if it is in the AM, “Good afternoon” if it is after 11:59am but before 5:00pm, “Good evening” if it is after 5:00pm but before 9:00pm, and just “Hello” (or something cheeky) otherwise. Add an optional argument, time, that has a default value of Sys.time() to help us check if our function works.

```{r}
print_greeting <- function(date = today(),user = Sys.info()[["user"]], time = Sys.time()){
  
  time_hour <- hour(time)
  
  if(time_hour < 12) greeting_start <- "Good morning,"
  else if(time_hour < 17) greeting_start <- "Good afternoon," 
  else if(time_hour < 21) greeting_start <- "Good evening,"
  else greeting_start <- "Hello"
  
  message <- paste0(greeting_start," ",user,"! Today is ", 
                    wday(date, abbr = FALSE, label = TRUE),
                    ",", month(date, label = TRUE, abbr = FALSE),
                    scales::ordinal(lubridate::day(today())),
                    ".")
  return(message)
}

print_greeting(user = "Florence Nightingale")
```

```{r}
# Time in morning
morning_time <- Sys.time()
hour(morning_time) <- 11

# Time in afternoon
afternoon_time <- Sys.time()
hour(afternoon_time) <- 16

# Time in evening
evening_time <- Sys.time()
hour(evening_time) <- 18

# Time at night
night_time <- Sys.time()
hour(night_time) <- 23

print_greeting(user = "Arjun",time = morning_time)
print_greeting(user = "Arjun",time = afternoon_time)
print_greeting(user = "Arjun",time = evening_time)
print_greeting(user = "Arjun",time = night_time)
```

