---
title: "Importing and Exporting Data"
author: "Arjun Mannem"
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
output:
  word_document: default
  html_document: default
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r }
library(data.table)
library(tidyverse)
library(arrow)
library(bench)
library(flextable)
library(ggbeeswarm)
```

### Importing data from outside R

Import the possum.csv data directly from the URL https://raw.githubusercontent.com/dilernia/STA418-518/main/Data/possum.csv into R using the read.csv() function creating an object called possums. Print the first 5 rows of the data using the slice_head() function and the first 6 columns using the select() function as in the code below.


```{r}
possums <- read.csv("https://raw.githubusercontent.com/dilernia/STA418-518/main/Data/possum.csv")

possums|>
  dplyr:: select(1:6) |>
  slice_head(n = 5) |>
  flextable() |>
  autofit()

```

Download the csv file we just imported, but then use the import wizard to import the possum.csv data using the read.csv() function as well.
 
*read as possums*
 
Download the parquet file from the URL https://raw.githubusercontent.com/dilernia/STA418-518/main/Data/possum.parquet and then import the data into R using the read_parquet() function from the arrow package using the code below. Display a subset of the data, only the first 5 rows of the data and the first 8 columns, and print the subset using the flextable() function.

```{r}
library(arrow)

# Downloading file from GitHub
download.file(url = "https://raw.githubusercontent.com/dilernia/STA418-518/main/Data/possum.parquet", destfile = "possum.parquet", mode = "wb")

# Import after manually downloading file
possums_parquet_data <- read_parquet("possum.parquet")
```

Import the possum.csv data directly from the URL https://raw.githubusercontent.com/dilernia/STA418-518/main/Data/possum.csv into R using the fread() function from the data.table package.

```{r}
possum_fread <- fread("https://raw.githubusercontent.com/dilernia/STA418-518/main/Data/possum.csv")
```

Let’s compare how quickly these functions import and export slightly larger data from R. 

First, we simulate data with 50,000 observations, 2 variables, and roughly 10% of its values being missing. 

```{r}
# Generating 'big data'
set.seed(1994)
x <- runif(5e4)
y <- runif(5e4)
x[sample(5e4, 5e3)] <- NA
y[sample(5e4, 5e3)] <- NA
bigData <- as.data.frame(x = x, y = y)

# Saving as CSV file w/ data.table
fwrite(bigData, "bigData.csv")

# Saving as parquet file
write_parquet(bigData, "bigData.parquet")

# Saving as RDS file
write_rds(bigData, "bigData.rds")
```

Reproduce the table and violin plots below comparing the differences in import speeds between the various functions using the code below.

```{r}
library(bench)

# Comparing run times
readBmResult <- mark(read.csv("bigData.csv"), read_csv("bigData.csv", show_col_types = FALSE), 
               fread("bigData.csv"), read_rds("bigData.rds"),
     read_parquet("bigData.parquet", as_tibble = TRUE),
     check = FALSE, min_iterations = 5) 

ggObj <- plot(readBmResult)

importTimes <- ggObj$data |> mutate(expression = paste0(map_chr(str_split(expression, pattern = "[(]"), 1), "()"))

# Printing table
importTimes |> 
  arrange(desc(median)) |> 
  dplyr::select(expression:mem_alloc) |> 
  distinct() |> 
  flextable() |> 
  autofit()

```

```{r}
# Creating violin plots
importTimes |> ggplot(aes(x = time, y = fct_reorder(expression, time))) + 
  geom_violin(fill = "dodgerblue") + 
  geom_jitter(height = 0.2, pch=21, fill = "black", color = "white") + 
  labs(title = "Comparison of read times", y = "", x = "Run time") + theme_bw()
```


### Exporting data from R

Reproduce the table and violin plots below comparing the differences in export speeds between the various functions using the code below.

```{r}
library(bench)

# Comparing run times
writeBmResult <- mark(write.csv(bigData,"bigData.csv"), write_csv(bigData,"bigData.csv"), 
               fwrite(bigData,"bigData.csv"), write_rds(bigData,"bigData.rds"),
     write_parquet(bigData,"bigData.parquet"),
     check = FALSE, min_iterations = 5) 

ggObj <- plot(writeBmResult)

exportTimes <- ggObj$data |> mutate(expression = paste0(map_chr(str_split(expression, pattern = "[(]"), 1), "()"))

# Printing table
exportTimes |> 
  arrange(desc(median)) |> 
  dplyr::select(expression:mem_alloc) |> 
  distinct() |> 
  flextable() |> 
  autofit()

```


```{r}
# Creating violin plots
exportTimes |> ggplot(aes(x = time, y = fct_reorder(expression, time))) + 
  geom_violin(fill = "dodgerblue") + 
  geom_jitter(height = 0.2, pch=21, fill = "black", color = "white") + 
  labs(title = "Comparison of write times", y = "", x = "Run time") + theme_bw()
```


