---
title: "Importing and Exporting"
author: "Mallikarjuna Mannem"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Let's load some packages
```{r}
library(data.table)
library(tidyverse)
library(knitr)
library(arrow)
library(bench)
library(ggbeeswarm)
library(haven)
```

#tidyverse is the main package we use in data manipulation and visualization
#data.table is used to organize data in R and import and export csv
#arrow work on different files of csv
#ggbeeswarm creating paths
#haven function allows data import and export from various files.
#knitr package to display tables

let's import and export some data in R using read.csv()

```{r}
#Importing possums data into R

possums <- read.csv("https://raw.githubusercontent.com/dilernia/STA418-518/main/Data/possum.csv")


library(readr)
possums <- read_csv("C:/Users/IT Loaner/Downloads/raw.githubusercontent.com_dilernia_STA418-518_main_Data_possum.csv")


possums %>% 
  dplyr::select(1:6) %>% 
  slice_head(n = 5) %>% 
  knitr::kable()
```

##Using `read_csv()`

```{r}
possums<- readr::read_csv("https://raw.githubusercontent.com/dilernia/STA418-518/main/Data/possum.csv")
```

```{r}
#Importing using R parquet
library(arrow)

# Import after manually downloading file
possums <- read_parquet("C:\\Users\\IT Loaner\\Downloads\\possum.parquet")
```

##Using `fread()` from the `data.table` package

```{r}
library(data.table)

# Import after manually downloading file
possums <- fread("https://raw.githubusercontent.com/dilernia/STA418-518/main/Data/possum.csv")
```




