---
title: "MLR_Homework"
author: "Arjun Mannem"
date: "`r Sys.Date()`"
output: word_document
---

### 1. First, add a new code chunk containing the code below which loads the tidyverse, ggthemes, and the flextable R packages.

```{r}
# Install packages if needed (note: this may take some time to run)
for(pack in c('tidyverse', 'ggthemes', 'flextable')) {
if(!(pack %in% installed.packages())) {
  install.packages(pack)
}
}

# Load necessary packages
library(tidyverse)
library(ggthemes)
library(flextable)
```

### 2. In another code chunk, load the data on hawks and set the default ggplot2 and flextable themes using the code below.

```{r}
# Set ggplot theme for visualizations
theme_set(ggthemes::theme_few())

# Create a counter for tables
tbl_counter <- 1

# Function for making flextable
make_flex <- function(myData, ndigits = 2, caption = NULL) {
  
  if("p.value" %in% colnames(myData)) {
    myData$p.value <- format.pval(myData$p.value, digits = ndigits)
  }
  
  myFlex <- myData %>% 
    flextable::flextable() %>% 
    flextable::autofit() %>% 
    flextable::colformat_double(digits = ndigits) %>% 
    flextable::fit_to_width(7.5)
  
  if(!is.null(caption)) {
        myFlex <- myFlex %>% 
      flextable::set_caption(paste0("Table ", tbl_counter, ": ", caption))
  }
  
  assign("tbl_counter", 
         value = tbl_counter + 1, envir = .GlobalEnv)
  
   return(myFlex) 
}

# Set options for flextables
set_flextable_defaults(na_str = "NA")

# Vectors of color-blind friendly colors
speciesColors <- c("#009E73", "#D55E00", "#56B4E9")
ageColors <- c("#CC79A7", "#E69F00", "#F0E442")

# Importing data on hawks
hawks <- read_csv("https://raw.githubusercontent.com/dilernia/STA323/main/Data/hawks.csv") %>% 
  dplyr::select(species, age, wing_mm:tail_mm)
```


### 3. Reproduce the scatter plot below showing the weight of the hawks (weight_g) by the tail lengths (tail_mm), and color the points based on the age category of each hawk using the code below.

```{r}
# Creating scatter plot coloring points by sex
hawks %>% 
  ggplot(aes(x = tail_mm, y = weight_g)) + 
  geom_point(aes(x = tail_mm, y = weight_g, fill = age), pch = 21, color = "white") + 
  labs(title = "Hawks Data: Weight by Tail Length",
       x = "Tail length (mm)",
       y = "Weight (grams)",
       fill = "Age") +
    scale_fill_manual(values = ageColors) +
  theme(legend.position = "bottom")
```


### 4. Modifying the code from the previous part, create a scatter plot showing the weight of the hawks (weight_g) by the wing span (wing_mm), and color the points based on the species of each hawk. Make sure to update all plot labels accordingly.

```{r}
# Creating scatter plot coloring points by sex
hawks %>% 
  ggplot(aes(x = wing_mm, y = weight_g)) + 
  geom_point(aes(x = wing_mm, y = weight_g, fill = species), pch = 21, color = "white") + 
  labs(title = "Hawks Data: Weight by Wing Span",
       x = "Wing span (mm)",
       y = "Weight (grams)",
       fill = "Species") +
    scale_fill_manual(values = speciesColors) +
  theme(legend.position = "bottom")
```

### 5. Using the lm() function in R, fit a multiple linear regression model with species (species) and wing span (wing_mm) as the predictor variables and weight (weight_g) as the response variable. Print the model coefficients and table of model summary metrics using the make_flex() function, reproducing the output below.

```{r}
# Fitting three-predictor multiple linear regression model
mlrModel <- lm(weight_g ~ species + wing_mm,
                data = hawks)

# Printing model output
mlrModel %>% 
  broom::tidy() %>% 
  make_flex(caption = "Coefficient estimates for model",
            ndigits = 2)
mlrModel %>% 
  broom::glance() %>% 
  make_flex(caption = "Model summary metrics",ndigits=2)
```

### 6. Provide a statement of the estimated regression equation using species (species) and wing span (wing_mm) as the predictor variables and weight (weight_g) as the response variable.


The estimated MLR model is

ŷ = -139.95 + 355.22 * x1 - 135.72 * x2 + 2.29 * x3,

where:

ŷ represents the estimated weight of a species in grams

βHat0 is considered as Coopers hawks

x1=1 for a red-tailed hawk, x1=0 otherwise

x2=1 => sharp-shinned hawk, x2=0 otherwise

x3 represents the wing span in mm


### 7. Provide the value of and interpret the estimated slope for wing span in context.

Since βHat3 = 2.29, for every 1 mm increase in wing span, we expect the weight of species to increase by 2.29 grams on average, holding the species constant.

### 8. Provide the value of and interpret the estimated slope for the indicator variable for red-tailed hawks in context.

Since βHat1 = 355.22, the expected weight in grams of a Red-tailed hawk species is 355.22 grams greater than that of Coopers hawk species on average, holding the wing span in mm constant.

### 9. Provide and interpret the value of r2 in this context.

r^2 = 0.90

*Interpretation*: About 90% of the variability in the weights of a species in grams is explained by the species and wing span in mm.


### 10. Obtain a predicted value and 90% prediction interval for a red-tailed hawk with a wing span of 400mm using the predict() function.

```{r}
# Obtaining predicted value
predict(mlrModel, newdata = tibble(species = "Red-tailed",wing_mm = 400),interval = "prediction",level=0.90)

```
*Predicted Value*: 1132.058 grams

*90% prediction interval* (in grams): (885.6923, 1378.424)

**Interpretation**: We expect 90% of Red-tailed hawks with wing spans of 400mm to have a body mass between 885.7 and 1378.4 grams.


### 11. Obtain a predicted value and 95% prediction interval for a Cooper’s hawk with a wing span of 250mm using the predict() function.

```{r}
# Obtaining predicted value
predict(mlrModel, newdata = tibble(species = "Coopers",wing_mm = 250),interval = "prediction",level=0.95)

```
*Predicted Value*: 433.0427 grams

*95% prediction interval* (in grams): (137.5703, 728.5152)

**Interpretation**: We expect 95% of Coopers hawks with wing spans of 250mm to have a body mass between 137.6 and 728.5 grams.

